---
title: "Introduction to Splines"
author: "Andy Shen, Devin Francom"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's say you want to fit a model using some wiggly data.  Maybe

```{r cars}
set.seed(12)
n<-300
x<-seq(0,1,length.out=n)
y<-sin(2*pi*x^2)*10+rnorm(n)
plot(x,y)
```

One way to fit a model to data like this is to come up with a linear basis and fit a linear model using the basis as the X matrix (which we will call B).  People often use splines as a basis.  The simplest set of spline basis functions would be to make the ith basis function (i.e., the ith column of B) look like
$$ B_{ij} = [s_i(x_j - t_i)]_+  $$
where $s\in \lbrace -1,1 \rbrace$, which we'll call the sign, and $t$ is a value in the domain of $x$, which we will call a knot.  Also, $[a]_+ = max(0,a)$.

Try some combinations of $s$ and $t$ to see what your basis functions look like, and what the corresponding linear model fit looks like (using the lm function or your Bayesian linear model code).  Try with different numbers of basis functions, also.

\pagebreak



# 3 Bayesian Spline with Random Knots

```{r, error=TRUE}
library(mvtnorm)

mcmc_spline <- function(nknot = 5) {
  
  p_t <- function(sig_sq, betahat, X) {
    (-1/(2*sig_sq)) * t(y - (X %*% betahat)) %*% (y - (X %*% betahat))
  }
  
  p_sig <- function(a = 0, b = 0, X, beta) {
    n <- nrow(X)
    a_term <- a + (n/2)
    b_term <- 0.5 * (2*b + (t(y - (X %*% beta)) %*% (y - (X %*% beta))))
    1 / rgamma(1, shape = a_term, rate = b_term)
  }
  
  p_beta <- function(sig_sq, tau_sq = 1000, X) {
    p = ncol(X)-1
    sig <- solve( (1/sig_sq) * (t(X) %*% X) + (1/tau_sq) * diag(p+1) )
    mu <- (1/sig_sq) * sig %*% t(X) %*% y
    rmvnorm(1, mean = mu, sigma = sig)
  }
  
  bhat <- function(sig_sq, tau_sq = 1000, X) {
    p = ncol(X)-1
    sig <- solve( (1/sig_sq) * (t(X) %*% X) + (1/tau_sq) * diag(p+1) )
    mu <- (1/sig_sq) * sig %*% t(X) %*% y
    mu
  }
  
  met_gibbs <- function(its) {
    
    mat_t <- matrix(NA, its, nknot)
    mat_t[1,] <- rep(0, nknot)
    
    ## INITIALIZE X_curr ##
    X_curr <- matrix(NA, nknot, length(x))
    s <- sample(c(1), nknot, replace = TRUE)
    for(i in 1:nknot) {
        for(j in 1:length(x)) {
          X_curr[i,j] <- max(s[i] * (x[j] - mat_t[1,i]), 0)
        } #creating basis
      } #X Matrix 
    X_curr <- t(X_curr); X_curr <- cbind(1, X_curr)
    
    mat_beta <- matrix(NA, its, ncol(X_curr))
    mat_sig <- rep(NA, its)
    
    mat_sig[1] <- 0.01
    mat_beta[1,] <- rep(0, ncol(X_curr))
    
    ar <- 0
    for(it in 2:its) {
      propose_t <- function(x) {
        log(dunif(x, 0, 1))
      }
    
      candidate_t <- runif(nknot, 0, 1)
      
      ### CREATE BASIS FUNCTIONS FROM THIS t-vector  ###
      
      s <- sample(c(1), nknot, replace = TRUE)
      
      X_cand <- matrix(NA, nknot, length(x))
      
      for(i in 1:nknot) {
        for(j in 1:length(x)) {
          X_cand[i,j] <- max(s[i] * (x[j] - candidate_t[i]), 0)
        } #creating basis
      } #X Matrix
      
      X_cand <- t(X_cand)
      X_cand <- cbind(1, X_cand)
      
      ratio <- 
        (p_t(sig_sq = mat_sig[it-1], betahat = bhat(mat_sig[it-1], X = X_cand), X = X_cand)
         - propose_t(candidate_t)) -
        (p_t(sig_sq = mat_sig[it-1], betahat = bhat(mat_sig[it-1], X = X_curr), X = X_curr)
         - propose_t(mat_t[it-1,]))
      
      accept_prob <- min(0, ratio)
      u <- log(runif(1))
      browser()
      if(u < accept_prob) {
        ar <- ar + 1
        mat_t[it,] <- candidate_t
        X_curr <- X_cand
      } else {
        mat_t[it,] <- mat_t[it-1,]
      }
      
      mat_beta[it,] <- p_beta(sig_sq = mat_sig[it-1], X = X_curr)
      mat_sig[it] <- p_sig(beta = mat_beta[it,], X = X_curr)
    }
    list(mat_beta, mat_sig, mat_t, X_curr, ar)
  }

  its <- 3000
  a <- met_gibbs(its = its)
  mat_beta <- a[[1]] #beta values
  mat_sig <- a[[2]] #sig^2 values
  mat_t <- a[[3]]
  X_curr <- a[[4]]
  ar <- a[[5]]
  
  list(mat_beta, mat_sig, mat_t, X_curr, ar)
}

```


```{r}
mcmc_spline()
```











# 1 Manual Frequentist Spline Function

```{r}
generate_spline <- function(tvec, y, x, nknot = length(tvec)) {
  s <- sample(c(1), nknot, replace = TRUE)
  Bmat <- matrix(NA, nknot, length(x))
  hs <- Bmat
  
  for(i in 1:nknot) {
    for(j in 1:length(x)) {
      Bmat[i,j] <- max(s[i] * (x[j] - tvec[i]), 0)
    } #creating basis functions
  }
  
  mBmat <- t(Bmat)
  mod <- lm(y ~ mBmat) #use gibbs to sample coefs in bayes
  pred <- predict(mod)
  
  sq <- x
  for(ii in 1:nknot) {
    if(s[i] == 1) {
      hs[ii,] <- sq - tvec[ii]
      hs[ii,][sq < tvec[ii]] <- 0
    }
    else {
      hs[ii,] <- -1*(sq - tvec[ii])
      hs[ii,][sq < tvec[ii]] <- 0
    }
  } #setting x values
  
  plot(x,y, main = "Manual Basis Spline")
  lines(x, pred, type = "l", lwd = 5, col="blue1")
  summary(mod)
}

```

```{r}
tv <- c(0, 0.525, 0.865) #vector of t-values
generate_spline(tv, y = y, x = x)
```

\pagebreak

# 2 Bayesian Spline

```{r}
library(mvtnorm)

bayes_spline <- function(tvec, y, x, nknot = length(tvec)) {
  s <- sample(c(1), nknot, replace = TRUE)
  Bmat <- matrix(NA, nknot, length(x))
  hs <- Bmat
  
  for(i in 1:nknot) {
    for(j in 1:length(x)) {
      Bmat[i,j] <- max(s[i] * (x[j] - tvec[i]), 0)
    } #creating basis
  } #X Matrix
  
  X <- t(Bmat)
  X <- cbind(1, X)
  
  p_sig <- function(a = 0, b = 0, n = nrow(X), beta) {
    a_term <- a + (n/2)
    b_term <- 0.5 * (2*b + (t(y - (X %*% beta)) %*% (y - (X %*% beta)))) #y, X defined above
    1 / rgamma(1, shape = a_term, rate = b_term)
  }
  
  p_beta <- function(sig_sq, tau_sq = 1000, p = ncol(X)-1) {
    sig <- solve( (1/sig_sq) * (t(X) %*% X) + (1/tau_sq) * diag(p+1) )
    mu <- (1/sig_sq) * sig %*% t(X) %*% y
    rmvnorm(1, mean = mu, sigma = sig)
  }
  
  gibbs <- function(its) {
    mat_beta <- matrix(NA, its, ncol(X))
    mat_sig <- rep(NA, its)
    mat_sig[1] <- 0.01
    mat_beta[1,] <- rep(0, ncol(X))
    for(it in 2:its) {
      mat_beta[it,] <- p_beta(sig_sq = mat_sig[it-1])
      mat_sig[it] <- p_sig(beta = mat_beta[it,])
    }
    list(mat_beta, mat_sig)
  }

  its <- 3000
  a <- gibbs(its = its)
  mat_beta <- a[[1]] #beta values
  mat_sig <- a[[2]] #sig^2 values  
  
  list(mat_beta, mat_sig, X)
}

bayesian <- bayes_spline(c(0, 0.525, 0.865), y = y, x = x)
mat_beta <- bayesian[[1]]
mat_sig <- bayesian[[2]]
X <- bayesian[[3]]
burn <- 1:15
colMeans(mat_beta[-burn,]) #target: -1.83, 25.19, -91.23, 135.03
mean(mat_sig[-burn]) #target: 1.46
```


```{r}
mean.pred <- X %*% t(mat_beta)
matplot(x, mean.pred, type = "l")
points(x,y)
```


```{r}
# pred <- t(mean.pred) + rnorm(300*2985,sd=sqrt(mat_sig)) #mean: X %*% beta
# hist(X[100,] %*% t(mat_beta) + rnorm(2985,sd=sqrt(mat_sig))) #prediction
# matplot(x,t(pred), type='l', col='lightgrey')
# matplot(x,mean.pred, type='l', col='green', add=T)
# points(x,y)
```

\pagebreak
